### 2018-11-08 treestats_paper.tex
notes from CSHL

\newcommand{\branch}{\mathcal{B}} % branch length stat
\newcommand{\site}{\mathcal{S}} % site stat

@@

@@

\plr{utility of looking at lots of individual-based stats}
With whole-genome sequence
we can confidently infer quantities
between \emph{individuals} that previously had to be estimated
averaging over populations.
This gives us much higher resolution, both in space and along the genome.
With geographic sampling,
there is potentially a lot of information available
from looking at all pairwise comparisons.

\plr{computational challenges}
Single statistics are not too bad to compute
since you can get them from allele frequencies,
but larger summaries, like the matrix of pairwise divergences between $N$ individuals,
naively involve $O(N^2)$ comparisons.

\plr{Something about how} in many applications
we'd really like to know the trees
and sequence differences really just stand in for these.
Example: $F_{ST}$ is X; 
$\theta_W$ is X;
divergence is X.

\plr{Review of phylogenetics lit}
a good review of theoretical properties \citet{semple_and_steel}

%%%%%%%%%%%%%%%%%%%%%%%
\section*{Methods}

We consider only so-called \emph{single site} statistics.
These are statistics computed as a simple \emph{average} 
of a quantity calcuated across a number of sites.
For a concrete example, to compute an element of the site frequency spectrum --
the proportion of sites at which an allele is present in 10\% of the samples, for instance --
we could 
(a) assign each site a value, either $+1$ if it has an allele at 10\%, or zero otherwise,
and
(b) take the average of these values across all sites.
As another example,
to compute the pairwise divergence between two sequences,
we would similarly assign each site either $+1$ or 0 according to whether the sequences differ,
and average these values.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}

%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Acknowledgement}

>>>>

\newcommand{\tskit}{{\texttt{tskit}}}
\newcommand{\branch}{\mathcal{B}} % branch stat
\newcommand{\site}{\mathcal{S}} % site stat
\newcommand{\node}{\mathcal{N}} % node stat

\newcommand{\nw}{{w}} % node weights
\newcommand{\iw}{{\tilde w}} % initial weights
\newcommand{\aw}{{\bar w}} % allele weights
\newcommand{\Tw}{{W}} % allele weights

@@

%%% OUTLINE
% 1. Framework and algorithms for computing general stats
% 
%     * Framework:
% 
%         - single-site stats are functions of the genotype vector
%         - in infinite sites, the genotype vector of a mutation is determined by the nodes below it
%         - ex: divergence as a sum over trees and branches
%         - ex: ancestry proportions
%         - general definition, using summary function of weights propagated up the tree
%             * recursion relating weights on nodes of two forests that differ by a branch
% 
%     * Review of tree sequences (quick)
% 
%     * Algorithms:
% 
%         - weight propagation across tree differences
%         - count by node length ("Node Statistics")
%         - count by branch area ("Branch Statistics")
%         - count by site ("Site Statistics")
%         - note on outputs (by node, by site, in windows)
% 
% 2. Examples: in sections, one of each type of output
% 
%     - ancestry proportions
%     - list of standard summary statistics: divergence, Fst, f4, covariance
%         * performance for one of these
%     - PCA: Krylov method
%         * performance
% 
% 3. Correspondence to tree statistics
% 
%     - math: in infinite sites, neutral model, E[site] = branch
%     - plot from simulation showing branch stats are less noisy versions of the site stats
%     - show time profile of PC1 as computed from rare variants and common variants

@@

\plr{need something different here}
Today's vast quantity of whole-genome sequence
makes it possible to confidently measure quantities
between \emph{individuals} that previously had to be roughly estimated
averaging over populations.
For instance, XXX estimated $F_{ST}$ between XX from allozyme sequence,
but now we know that $F_{ST}$ can be viewed as $1 - t_{within}/t_{total}$ \citep{slatkin_fst}
which can therefore be estimated precisely by $1 - \pi_{\within}/\pi_{total}$.

% {computational challenges}
Computation is beginning to be a major problem:
whole-genome sequence on the scale of the UK Biobank,
for instance, might have 500K individuals genotyped at 10M loci (??),
resulting in a genotype matrix of $5 \times 10^{12}$ entries.
Computing a single summary from this matrix is still relatively quick,
once loaded into memory,
but for many applications we are now interested in computing a large number of statistics
(for instance, consider the roughly $10^{11}$ different possible pairwise divergences).
Indeed, many sorts of modern inference machinery, such as deep learning,
depend on being able to compute enough statistics that this becomes a serious bottleneck in practice.

The \emph{succinct tree sequence} is a data structure that efficiently encodes
the genealogical trees that describe how a set of individuals are related to each other
at each point along their genome.
The data structure was introduced for coalescent simulation by \citet{kelleher2016efficient},
and refined for forwards simulation by \citet{kelleher2018efficient}.
These genealogical trees change as one looks across the genome
because of the results of recombination in ancestors.
However, nearby genealogical trees tend to share a lot of common structure,
and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
are often shared across relatively long distances,
which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
\citet{kelleher2016efficient}
showed how this this redundancy can be used
to both store genome sequence and the associated genealogical relationships extremely compactly,
as well as to compute statistics extremely efficiently.
This paper generalizes the latter strategy
to a much broader class of statistics.

\plr{Review of tree shape stuff in phylogenetics}
a good review of theoretical properties \citet{semple_and_steel}.

% summary
In this paper, we present a general framework for defining and efficiently computing
``single-site'' genetic statistics,
i.e., statistics of aligned genome sequence that can be expressed as averages over values computed
separately for each site.
Patterns of genetic polymorphism in a population 
are determined by the population's history of Mendelian inheritance,
as summarized by the underlying genealogical trees,
and so it is natural (and helpful) to express these statistics
in terms of the structure of the trees themselves.
This perspective turns out to both suggest efficient algorithms 
for computing these statistics on tree sequences,
and allows us to treat single-site genetic statistics in a more general framework
of summaries of tree shape and inheritance.
The methods are implemented in the python package \tskit.

%%%%%%%%%%%%%%%%%%%%%%%
\section*{Framework and algorithms}

In this paper, 
we consider only so-called \emph{single site} statistics
(which we will define concretely shortly).
Extensions to statistics involving the pairwise joint distribution of genotypes across sites,
such as linkage disequlibrium,
are straightforward, and are planned for future work.
Haplotype-based statistics may require a different class of algorithms.

%%%%%%%%%%%%%%%%%%
\subsection*{Theoretical framework}

Suppose we have genotyped a collection of $N$ (homologous) chromosomes
at a sequence of $L$ loci along the genome.
Suppose first that we wish to compute an additive function of the allele frequency spectrum.
In other words, suppose that the frequency of allele $a$ at a given site is $p_i(a)$,
and that the allele frequency spectrum 
-- i.e., the proportion of alleles whose frequency is $p$ in our sample -- is $f(p)$,
and for some function $h()$ we want to compute $\sum_p f(p) h(p)$.
Since this is a simple average over sites,
we can compute this, summing over the $L$ sites, as 
$\sum_p f(p) h(p) = (1/L) \sum_{i=1}^L \sum_a h(p_i(a))$.
This definition applies to multiallelic sites by assuming that for each allele
we can only see how many samples carry that allele,
not how the remaining samples are partitioned between one or more other alleles.
(Note also that this does not differentiate ancestral and derived alleles,
and so is a function of the \emph{folded} allele frequency spectrum.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
occur below that mutation in the tree (and removing those that subsequently mutate).

There are $2^n$ possible genotype patterns for $n$ haploid samples genotyped at a biallelic locus,
and so a general statistic of biallelic genotypes could be represented by a vector of length $2^n$.
A more convenient way to parameterize many functions of genotype patterns
uses three ingredients:
\begin{itemize}

    \item[(a)] A list of $K$ vectors of \emph{initial node weights}, $(\iw_1, \ldots, \iw_k)$, 
        where $\iw_i(j)$ is the ``initial'' value of the $i^\text{th}$ weight assigned to node $j$.

    \item[(b)] The \emph{weight} $\nw$ of a node in a tree
        is a $k$-vector of weights is $\nw = (\nw_1, \ldots, \nw_k)$,
        and is equal to the sum of initial weights of all nodes below it in the tree
        (including itself).

    \item[(c)] A \emph{summary function} $F : \R^k \to \R$ so that 
        $F(\nw)$ is the ``summary'' statistic we compute

\end{itemize}
Roughly speaking, we use these to compute a statistic
by first propagating the initial weights up the trees,
and then evaluate the summary function on the resulting vectors of weights
for each allele.

% Sum weights for each allele.
Concretely, to compute the statistic at a given site $j$,
for each allele $a$ seen at this site,
let $\aw_j(a)$, the weight associated with this allele,
be the sum of the initial weights over all nodes who inherit allele $a$.
If the site has only one mutation, which produced allele $a$ 
at node $i$, then $\aw_j(a) = \nw(i)$.
Then, the statistic computed at site $j$ is
\begin{align}
    \site(F, \iw)_j
    &=
    \sum_a \aw_j(a) ,
\end{align}
and a summary statistic across $L$ sites is
\begin{align}
    \bar \site(F, \iw)
    &=
    \frac{1}{L} \sum_{j=1}^L \sum_a \aw_j(a) .
\end{align}
We call this the \textbf{site statistic} associated with $\iw$ and $F$.

\plr{Figure: example of propagating weights and labeling nodes with weights,
    in the case that weights are number of samples you are ancestor of.
    Also, this is a node statistic.}

\plr{Figure: example of branch and site statistics}

This same framework can be used to compute other properties of interest.
For instance, suppse that we'd like to compute
what proportion of the genomes of the final generation were contributed by each ancestor.
In the same framework we can do ancestry proportions.
\plr{EXPLAIN, DEFINE NODE STATS}

\plr{Introduce branch stats.}
The \textbf{branch statistic} associated with the branch of length $t$ 
above node $n$ on a tree of length $\ell$,
with total weight $\Tw = \sum_j \iw(j)$
and node weights $\nw$,
is
\begin{align}
    \branch(F, \iw)_n
    &=
    \ell t ( F(\Tw - \nw(n)) + F(\nw(n)) ) .
\end{align}
This is the summary value that would be added to a site statistic 
if a single mutation occurred on this branch:
we must apply the summary function 
to both the weight below $n$ as well as the weight \emph{not} below $n$
to account for both derived and ancestral alleles.

\paragraph{Summary}
We have defined three types of statistic using the same weighting scheme:
\begin{enumerate}
    \item \textbf{Site statistics} 
        -- summarizes weights below each allele at each polymorphic site.
    \item \textbf{Node statistics} 
        -- summarizes total weight below each node.
    \item \textbf{Branch statistics}
        -- summarizes total weight below each branch, multiplied by branch length.
\end{enumerate}

\plr{TO-DO:} 
\begin{itemize}

    \item How will polarization work? 
        Right now, Site and Branch stats are polarized but Node statistics are not.
        (To make Node statistics unpolarized, we need to also add $F(\Tw - w)$ 
        to the weight of each node.)
        Perhaps this is OK -- Site and Branch stats should correspond to each other,
        but Node statistics are different, so could have different behavior.
        BUT, people will need to compute polarized Site and Branch stats;
        we need to make it easy for them somewhere.

    \item What about normalization?
        In the initial implementation, Node statistics were divided by the length
        for which the node was an ancestor to any sample, rather than just total sequence length
        To make these more analogous to the other stats, we think we want to remove this.
        The denominator for the original case is easily computable, by using 
        $F(x) = 1$ if $x>0$ and $F(x)=0$ otherwise, with appropriate (positive) weights.

\end{itemize}

@@

%%%%%%%%%%%%%%%%%%
\subsection*{Algorithms for fast computation}

%%%%%%%%%%%%%%%%%%
\section*{Case studies}
% AKA "examples"

%%%%%%%%%%%%%%%%%%
\subsection*{Ancestry proportions}

%%%%%%%%%%%%%%%%%%
\subsection*{Population genetics statistics}

%%%%%%%%%%%%%%%%%%
\subsection*{Genomic PCA}

%%%%%%%%%%%%%%%%%%
\section*{Duality of site and branch statistics}

Under a neutral, infinite-sites model of mutation with constant mutation rate across time,
the expected number of mutations per branch is proportional to its length.
This implies an isormorphism between ``site'' and ``branch'' statistics defined above.

%%%%%%%%%%%%%%%%%%
\subsection*{The duality}

%%%%%%%%%%%%%%%%%%
\subsection*{Separating mutational from genealogical noise}

%%%%%%%%%%%%%%%%%%
\subsection*{Visualizing the time profile of PCA}

\paragraph{Diploid statistics}
The statistics we develop here are in essence \emph{haploid} statistics.
These tools can be used to compute a statistic about polyploid individuals,
but they do not scale to large numbers of individuals.
For instance, to compute the heterozygosity of a set of samples
-- this is $(1/N) \sum_{i=1}^N 2 p_i (1-p_i)$, where $p_i$ is the derived allele frequency
in the $i^\text{th}$ individual --
one would need to have an entry in the weight vector for \emph{each individual}.
The genotype of a diploid at a site with a single mutation that occurred on the branch above node $n$
is determined by where that node falls relative to the diploid's two haplotypes in the tree:
if $n$ is on the path from either haplotype to their MRCA, the individual is heterozygous;
if $n$ is on the path from the MRCA to the root, the individual is homozygous derived;
and the individual is homozygous ancestral otherwise.
This suggests the following generalization of our current scheme for diploids:
for a list $(I_1, \ldots, I_N)$ of pairs of nodes,
and two vectors $w^{(1)}$ and $w^{(2)}$ of weights of length $N$,
add weight $w^{(1)}_i$ to each node on the path between the two nodes of $I_i$,
and weight $w^{(2)}_i$ to each node on the path from the MRCA of $I_i$ to the root.
With $k$ such weighting schemes, we would then need the summary function $F$ 
to take $k \times 2$ values as input.
Efficiently updating these ``diploid'' weights on the nodes should be possible,
but is not a straightforward extension of our current method.

%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Acknowledgements}

\newpage
\appendix

{\Large \plr{OLD STUFF}}

@@
